Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2015-04-13 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Clean up bib errors and latex file

In the United States, the loss of life due to unwanted fire has been decreasing steadily since the 1970’s.  This trend does not include loss of life as a result of the terrorist attack on September 11, 2001.  The estimated number of fire deaths in the U.S. in 1975 was 8,100 or 3.75 deaths per 100,000 population~\cite{America_Burning_Revisited}.  In 2013, there were 3,240 civilian fire fatalities or 1.02 deaths per 100,000 population~\cite{Karter:2014}.  During this same period, the number of fires in the United States has decreased from 2.5 million fires to 1.24 million fires.  There are many reasons that led to this decrease in loss of life due to fire in the United States.  An example would be research and data that enabled the development of product safety standards, improved building codes, improved fire safety systems, and fire prevention programs to name just a few.  None of this happened without the constant effort of agencies at the Federal, state, and, local levels working with professional fire organizations to deliver the five E’s of fire prevention; Engineering, Education, Enforcement, Economic Incentive, and Emergency Response~\cite{FIVE}.

@@

Polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishing and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Touch}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture and mattresses and bedding were the first items ignited in 33~\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

@@

Based on the field data developed using Streams average flame heights for each fuel were determined.  In Table~\ref{tab:Average_Flame_Heights}, each of the values represents the average flame heights, given in mm with 95~\% confidence limits based on a Type A statistical analysis of the data, for each of the source fuels shown as the percentage of time the flame was visible at that given height.  For the natural gas fires, the averages are from ten 30 s segments, while the fire was monitored at approximately 80~kW.  In some cases more than one 30 s segment was sampled per heat release rate experiment.  Sixteen 30~s video segments, which included the peak heat release rate, were successfully analyzed from the seventeen gasoline heat release rate experiments.  One of the videos was unusable for analysis due to reflected light which could not be edited out of the images.  For same reason, the polyurethane foam flame height averages are the result from eight of the ten heat release rate experiments.

@@

Figures~\ref{TC_Plume_TWNG_comp} through \ref{TC_Back_Center_TWNG_comp} show a complete set of data, averaged over two replicates, for the instrumented, non combustible wall natural gas burner experiments.  Comparing the temperatures of the plume at the most distant point from the wall, 0.61~m and with the plume temperatures with the burner contacting the wall in Figure~\ref{TC_Plume_TWNG_comp} shows an increase.  However for the burner position, 0.15~m from the wall, a decrease in plume temperatures occurs after approximately 3 minutes after ignition.  A review of the heat release rates and the test videos do not yield any clues as to the reason for this decrease which occurred in both experiments.  In both experiments the heat release rate was steady and the video did not show any change in flame movement.
Figures~\ref{TC_Surf_Cent_TWNG_comp} and \ref{TC_Surf_Edge_TWNG_comp} show consistent trends of the thermocouple temperatures in the vertical arrays on the wall increasing as the burner was moved toward the wall.  Figure~\ref{TC_Back_Center_TWNG_comp} shows that the peak temperatures aligned with the vertical center of the burner on the backside of the gypsum wallboard peaked at approximately 100~$^\circ$C.

@@

These instrumented non-combustible wall experiments provide a data set for the fires conducted that can be compared against ignition data for gypsum wallboard from the literature to determine the potential for ignition of the paper surface of gypsum wallboard at a given distance from the wall with a given fuel. The Ignition Handbook by Babrauskas is the most comprehensive set of ignition data and related information ever written~\cite{Babrauskas:2003}.  Table 19 in the Ignition Handbook includes calculated ignition temperatures from experimental data from two separate sources.

@@

Figures~\ref{IWGB_HF_Comp_Center} and \ref{IWGB_HF_Comp_Edge} present the averaged peak values from the two heat flux gauge arrays.  For the gasoline fire experiment at the 0.08~m position the wall temperatures and the heat flux values from the burner edge array had values that were higher than when those measured at the 0.0 m position. The peak heat release rates for the positions at 0.0~m, 0.08~m and 0.15~m were 120~kW, 110~kW, and 110~kW respectively.  So it does not seem that a change in HRR was the source of the increase.

@@

As defined in NFPA 921, a fire is the visible or measurable physical changes, or identifiable shapes, formed by a fire effect or group of fire effects~\cite{NFPA_921}.  In order to measure the dimensions of the fire pattern, such as height and width, a determination had to be made as to where the fire pattern stops.  The boundary between the undamaged area and the area that has been damaged by the fire is known as the line of demarcation~\cite{nfpa_921}.  As presented in the introduction, DeHaan~\cite{DeHaan:2012} categorizes the fire effects that form patterns as follows:
\begin{enumerate}
	\item Surface deposits -– no irreversible effect on surface
	\item Surface thermal effects –- physical change such a discoloration or melting
	\item Charring -– evidence of surface burning
	\item Penetration -– charring below the surface
	\item Consumption –- loss of surface material, charring throughout
\end{enumerate}

@@

 The initial compartment experimental series were conducted to examine the impact of the method of construction of the walls on the development of the fire patterns. These experiments would determine how the walls for the remaining compartment fire pattern experiments would be constructed.  Three cases were examined: 1) Open back, 12.7~mm thick, regular core, painted gypsum board on the front of a wood frame only, 2) Closed back, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame representative of an interior wall, and 3)	Insulated, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame, with insulation in the voids of the wood frame representative of an exterior wall. The insulation was 90~mm thick fiber glass batting,a kraft paper facing, and an "R value" of 13.  From the instrumented gypsum wallboard experiments, the peak temperatures on the backside of the wallboard occurred when the burners were positioned against the wall.  The peak temperatures were similar for all of the fuels, approximately 80~$^\circ$C, although the duration of the higher temperatures on the rear of the wallboard was longer for the natural gas and the gasoline fueled burners. Therefore the wall construction experiments were only conducted with the natural gas and gasoline fueled burners as the impact of insulating the back of the walls exposed to the fuels with the highest energy would produce the most significant differences, if any, to the fire patterns.

@@

The burner was centered on the middle panel of the rear wall and was in contact with the wall. The wall behind the center panel of the rear wall was constructed either as open, closed, or insulated to meet the needs of the experiment. The burner was positioned against the face of the painted gypsum board. As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment. The source fires remained the same in terms of fuels, amounts of fuels used, burner size and burn duration.  The natural gas had a 300 second burn time which was approximately the upper bound for the total burn time for the 500~mL of gasoline.  By using the same burners, these compartment fire pattern experiments can be correlated with the instrumented wall experimental results.

@@

Figure~\ref{NG_Open_Wall} shows a comparison of the ten fire patterns generated by the natural gas burner on the wall with open back construction.  The images within the figure are slightly different sized however they are scaled so that the images are comparable.        
The grids were used to assist with the determination of the fire pattern area and the sizing of the images. In order to measure the dimensions of the fire pattern, such as height and width, a determination had to be made as to where the fire pattern stops.  The boundary between the undamaged area and the area that has been damaged by the fire is known as the line of demarcation~\cite{nfpa_921}.

>>>>

In the United States, the loss of life due to unwanted fire has been decreasing steadily since the 1970’s.  This trend does not include loss of life as a result of the terrorist attack on September 11, 2001.  The estimated number of fire deaths in the U.S. in 1975 was 8,100 or 3.75 deaths per 100,000 population~\cite{America_Burning_Revisited}.  In 2013, there were 3,240 civilian fire fatalities or 1.02 deaths per 100,000 population~\cite{Karter:2014}.  During this same period, the number of fires in the United States has decreased from 2.5 million fires to 1.24 million fires.  There are many reasons that led to this decrease in loss of life due to fire in the United States.  An example would be research and data that enabled the development of product safety standards, improved building codes, improved fire safety systems, and fire prevention programs to name just a few.  None of this happened without the constant effort of agencies at the Federal, state, and, local levels working with professional fire organizations to deliver the five E’s of fire prevention; Engineering, Education, Enforcement, Economic Incentive, and Emergency Response~\cite{FEMA:2013}.

@@

Polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishing and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture and mattresses and bedding were the first items ignited in 33~\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

@@

Based on the field data developed using Streams average flame heights for each fuel were determined.  In Table~\ref{tab:Average_Flame_Heights}, each of the values represents the average flame heights, given in mm with 95~\% confidence limits based on a Type A statistical analysis of the data, for each of the source fuels shown as the percentage of time the flame was visible at that given height.  For the natural gas fires, the averages are from ten 30~s segments, while the fire was monitored at approximately 80~kW.  In some cases more than one 30~s segment was sampled per heat release rate experiment.  Sixteen 30~s video segments, which included the peak heat release rate, were successfully analyzed from the seventeen gasoline heat release rate experiments.  One of the videos was unusable for analysis due to reflected light which could not be edited out of the images.  For same reason, the polyurethane foam flame height averages are the result from eight of the ten heat release rate experiments.

@@

Figures~\ref{TC_Plume_TWNG_comp} through \ref{TC_Back_Center_TWNG_comp} show a complete set of data, averaged over two replicates, for the instrumented, non combustible wall natural gas burner experiments.  Comparing the temperatures of the plume at the most distant point from the wall, 0.61~m and with the plume temperatures with the burner contacting the wall in Figure~\ref{TC_Plume_TWNG_comp} shows an increase.  However for the burner position, 0.15~m from the wall, a decrease in plume temperatures occurs after approximately 3~min after ignition.  A review of the heat release rates and the test videos do not yield any clues as to the reason for this decrease which occurred in both experiments.  In both experiments the heat release rate was steady and the video did not show any change in flame movement.
Figures~\ref{TC_Surf_Cent_TWNG_comp} and \ref{TC_Surf_Edge_TWNG_comp} show consistent trends of the thermocouple temperatures in the vertical arrays on the wall increasing as the burner was moved toward the wall.  Figure~\ref{TC_Back_Center_TWNG_comp} shows that the peak temperatures aligned with the vertical center of the burner on the backside of the gypsum wallboard peaked at approximately 100~$^\circ$C.

@@

These instrumented non-combustible wall experiments provide a data set for the fires conducted that can be compared against ignition data for gypsum wallboard from the literature to determine the potential for ignition of the paper surface of gypsum wallboard at a given distance from the wall with a given fuel. The Ignition Handbook by Babrauskas is the most comprehensive set of ignition data and related information ever written~\cite{Babrauskas:2003}.  Table~19 in the Ignition Handbook includes calculated ignition temperatures from experimental data from two separate sources.

@@

Figures~\ref{IWGB_HF_Comp_Center} and \ref{IWGB_HF_Comp_Edge} present the averaged peak values from the two heat flux gauge arrays.  For the gasoline fire experiment at the 0.08~m position the wall temperatures and the heat flux values from the burner edge array had values that were higher than when those measured at the 0.0~m position. The peak heat release rates for the positions at 0.0~m, 0.08~m and 0.15~m were 120~kW, 110~kW, and 110~kW respectively.  So it does not seem that a change in HRR was the source of the increase.

@@

As defined in NFPA 921, a fire is the visible or measurable physical changes, or identifiable shapes, formed by a fire effect or group of fire effects~\cite{NFPA_921}.  In order to measure the dimensions of the fire pattern, such as height and width, a determination had to be made as to where the fire pattern stops.  The boundary between the undamaged area and the area that has been damaged by the fire is known as the line of demarcation~\cite{NFPA_921}.  As presented in the introduction, DeHaan~\cite{DeHaan:2012} categorizes the fire effects that form patterns as follows:
\begin{enumerate}
	\item Surface deposits -– no irreversible effect on surface
	\item Surface thermal effects –- physical change such a discoloration or melting
	\item Charring -– evidence of surface burning
	\item Penetration -– charring below the surface
	\item Consumption –- loss of surface material, charring throughout
\end{enumerate}

@@

 The initial compartment experimental series were conducted to examine the impact of the method of construction of the walls on the development of the fire patterns. These experiments would determine how the walls for the remaining compartment fire pattern experiments would be constructed.  Three cases were examined: 1) Open back, 12.7~mm thick, regular core, painted gypsum board on the front of a wood frame only, 2) Closed back, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame representative of an interior wall, and 3)	Insulated, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame, with insulation in the voids of the wood frame representative of an exterior wall. The insulation was 90~mm thick fiber glass batting,a kraft paper facing, and an ``R value'' of 13.  From the instrumented gypsum wallboard experiments, the peak temperatures on the backside of the wallboard occurred when the burners were positioned against the wall.  The peak temperatures were similar for all of the fuels, approximately 80~$^\circ$C, although the duration of the higher temperatures on the rear of the wallboard was longer for the natural gas and the gasoline fueled burners. Therefore the wall construction experiments were only conducted with the natural gas and gasoline fueled burners as the impact of insulating the back of the walls exposed to the fuels with the highest energy would produce the most significant differences, if any, to the fire patterns.

@@

The burner was centered on the middle panel of the rear wall and was in contact with the wall. The wall behind the center panel of the rear wall was constructed either as open, closed, or insulated to meet the needs of the experiment. The burner was positioned against the face of the painted gypsum board. As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment. The source fires remained the same in terms of fuels, amounts of fuels used, burner size and burn duration.  The natural gas had a 300~s burn time which was approximately the upper bound for the total burn time for the 500~mL of gasoline.  By using the same burners, these compartment fire pattern experiments can be correlated with the instrumented wall experimental results.

@@

Figure~\ref{NG_Open_Wall} shows a comparison of the ten fire patterns generated by the natural gas burner on the wall with open back construction.  The images within the figure are slightly different sized however they are scaled so that the images are comparable.
The grids were used to assist with the determination of the fire pattern area and the sizing of the images. In order to measure the dimensions of the fire pattern, such as height and width, a determination had to be made as to where the fire pattern stops.  The boundary between the undamaged area and the area that has been damaged by the fire is known as the line of demarcation~\cite{NFPA_921}.
